Soling at the 2000 Summer Olympics

### 2020-07-23 inst/f1points.Rmd
more consistentification

where $p_H$ is a quantification of the beneficial home ground effect.
Similar techniques have been used to account for the first-move
advantage in chess, and effective coordination between members of
doubles tennis teams; we may use a similar device to account for pole
position in Formula 1.  Here I analyse racing results from the 2017
season using the `hyper2` package \citep{hankin2017} which implements
the Plackett-Luce likelihood function with additional reified
entities.

@@

Here we use the results from 2017:

@@

We see in figure \ref{piechartstrength} that the driver with the
largest estimated strength was Rosberg in 2016 (at about 49\%) and in
years 2017-8-9 was Hamilton (at about 29\% 35\% 50\ respectively).
However, it should be noted that a likelihood ratio test
[`samep.test()`, supplied with the package] does not reject, for
example, the null that Hamilton and Vettel have the same strength in
2017 ($H_0\colon \pham= \pvet$) with a likelihood ratio of
$e^{1.5}\simeq 4.48$, corresponding (by Wilks's theorem) to an
asymptotic $p$-value of about $0.08$.

@@

```{r, label=calcpoints, echo=FALSE, cache=TRUE}
p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
points_real <- p
points_inaugural <- c(8,6,4,3,2)
dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
dp <- sort(dp,decreasing=TRUE)
dp[] <- seq_along(dp)
m <- maxp(H2017)
m <- sort(m,decreasing=TRUE)
m[] <- seq_along(m)
```

@@

```{r,label=likevspoints,cache=TRUE,echo=FALSE,fig.cap="Ordering of the \\label{orderingbypointsandlikelihood} top-ranked 11 drivers in the 2017 Formula 1 season.  Horizontal axis gives official (points-based) order, and the vertical axis gives the likelihood horder.    Thus the points-based ordering would be Hamilton, Vettel, R\"{a}ikk\"{o}nen YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE while the likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas"}
par(pty='s')        # square plot
ox <- dp
oy <- ordertrans(m,dp)
plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
    xlab="official order",ylab="my order",main='Formula 1, 2017 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

From figure \ref{orderingbypointsandlikelihood}, we see that the
first two drivers (Hamilton, Vettel) are are respectively first and
second according to both ranking ranking procedures; but
R\"{a}ikk\"{o}nen YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE and
Bottas are third and fourth, and fourth and third, according to points
and likelihood respectively.  The degree of agreement between the two
systems might be quantified as "two", for the positions of the top two
drivers agree.

Observing that both the likelihood ranking and the points ranking are
random variables suggests a method whereby we can objectively assess a
given points system.  Using sampling techniques we can repeatedly
generate an order table \emph{in silico}, using estimated driver
strengths from the observed 2017 table.  For each of, say, 1000 such
synthetic tables, calculate drivers' maximum likelihood
\citeauthor{plackett1975} strengths, and also their points awarded
according to any given points system.  We then compare rankings
generated by the \citeauthor{plackett1975} strengths and the points
awarded and note the degree of agreement between the two.  This
furnishes an objective assessment of the points system used.

@@

(this happens to be identical to the ranking after the extra point was
awarded for fastest lap).  However, if we were to use Zipf's law, the
ranking would be:

```{r,label=zipfpointsshower,echo=FALSE}
zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),1/seq_len(top))
zipf <- round(zipf[seq_len(top)],2)
zipf <- sort(zipf,decreasing=TRUE)
zipf[] <- seq_along(zipf)
zipf
```

@@

Now compare the likelihood ranking:

@@

>>>>

where $p_H$ is a quantification of the beneficial home ground effect.
Similar techniques have been used to account for the first-move
advantage in chess, and effective coordination between members of
doubles tennis teams; we may use a similar device to account for pole
position in Formula 1.  Here I analyse seasons 2016-2019 using the
`hyper2` package \citep{hankin2017} which implements the Plackett-Luce
likelihood function with additional reified entities.

@@

Taking 2017 as an example, we see the rank table as follows:

@@

We see in figure \ref{piechartstrength} that the driver with the
largest estimated strength was Rosberg in 2016 (at about 49\%) and in
years 2017-8-9 was Hamilton (at about 29\%, 35\%, and 50\%
respectively).  However, it should be noted that a likelihood ratio
test [`samep.test()`, supplied with the package] does not reject, for
example, the null that Hamilton and Vettel have the same strength in
2017 ($H_0\colon \pham= \pvet$) with a likelihood ratio of
$e^{1.5}\simeq 4.48$, corresponding (by Wilks's theorem) to an
asymptotic $p$-value of about $0.08$.

@@

```{r, label=calcpoints, echo=FALSE, cache=TRUE}
p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
points_real <- p
points_inaugural <- c(8,6,4,3,2)
dp2016 <- ordertable2points(as.ordertable(F1_table_2016[seq_len(top),]),p)
dp2016 <- sort(dp2016,decreasing=TRUE)
dp2016[] <- seq_along(dp2016)
m2016 <- maxp(H2016)
m2016 <- sort(m2016,decreasing=TRUE)
m2016[] <- seq_along(m2016)

dp2017 <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
dp2017 <- sort(dp2017,decreasing=TRUE)
dp2017[] <- seq_along(dp2017)
m2017 <- maxp(H2017)
m2017 <- sort(m2017,decreasing=TRUE)
m2017[] <- seq_along(m2017)

dp2018 <- ordertable2points(as.ordertable(F1_table_2018[seq_len(top),]),p)
dp2018 <- sort(dp2018,decreasing=TRUE)
dp2018[] <- seq_along(dp2018)
m2018 <- maxp(H2018)
m2018 <- sort(m2018,decreasing=TRUE)
m2018[] <- seq_along(m2018)

dp2019 <- ordertable2points(as.ordertable(F1_table_2019[seq_len(top),]),p)
dp2019 <- sort(dp2019,decreasing=TRUE)
dp2019[] <- seq_along(dp2019)
m2019 <- maxp(H2019)
m2019 <- sort(m2019,decreasing=TRUE)
m2019[] <- seq_along(m2019)
```

@@

```{r,label=likevspoints,cache=TRUE,echo=FALSE,fig.cap="Ordering of the \\label{orderingbypointsandlikelihood} top-ranked 11 drivers in the 2017 Formula 1 season.  Horizontal axis gives official (points-based) order, and the vertical axis gives the likelihood horder.    Thus, taking 201 as an example,  the points-based ordering would be Hamilton first, then Vettel, then R\"{a}ikk\"{o}nen; YES I KNOW IT IS WRONG: RMARKDOWN IS DEFECTIVE while the likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas ALSO R PLOTTING SUCKS HERE AND I WILL HAVE TO CHANGE THE PDF BY HAND WITH INKSCAPE TO MAKE IT LOOK GOOD"}
par(mfrow=c(2,2))
par(mar = c(3,3,3,3))

par(pty='s')        # square plot
ox <- dp2016
oy <- ordertrans(m2016,dp2016)
plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
    xlab="points order",ylab="likelihood order",main='2016 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)

par(pty='s')        # square plot
ox <- dp2017
oy <- ordertrans(m2017,dp2017)
plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
    xlab="points order",ylab="likelihood order",main='2017 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray', cex=0.6) }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)

par(pty='s')        # square plot
ox <- dp2018
oy <- ordertrans(m2018,dp2018)
plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
    xlab="points order",ylab="likelihood order",main='2018 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray', cex=0.6) }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)

par(pty='s')        # square plot
ox <- dp2019
oy <- ordertrans(m2019,dp2019)
plot(ox,oy,asp=1,pty='s',xlim=c(0,11.2),ylim=c(0,11.2),pch=16,
    xlab="points order",ylab="likelihood order",main='2019 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray',cex=0.6) }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

From figure \ref{orderingbypointsandlikelihood}, we see a comparison
between the two ordering systems.  Taking 2017 as an example, drivers
Hamilton and Vettel are are respectively first and second according to
both ranking ranking procedures; but R\"{a}ikk\"{o}nen YES I KNOW IT
IS WRONG: RMARKDOWN IS DEFECTIVE and Bottas are third and fourth, and
fourth and third, according to points and likelihood respectively.
The degree of agreement between the two systems might be quantified as
"2", as the positions of the top two drivers agree.  Thus the degree
of agreement for the years 2016-9 would be 8,2,6, and 2 respectively.

@@

(this happens to be identical to the ranking after the extra point was
awarded for fastest lap).  However, if we were award points according
to Zipf's law we would have:

```{r,label=zipfpointsshower,echo=FALSE}
zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),1)
zipf <- round(zipf[seq_len(top)],2)
zipf <- sort(zipf,decreasing=TRUE)
zipf[] <- seq_along(zipf)
zipf
```

@@

@@

Thus, for 2017 at least, we see that the current poionts system agrees
with likelihood ranking for the top four places, while Zipf's law
agrees to three.

Observing that both the likelihood ranking and the points ranking are
random variables in this paradigm suggests a method whereby we can
objectively assess a given points system.  Using sampling techniques
we can repeatedly generate an order table \emph{in silico}, using
estimated driver strengths from the observed tables.  For each of,
say, 1000 such synthetic tables, calculate drivers' maximum likelihood
\citeauthor{plackett1975} strengths, and also their points awarded
according to any given points system.  We then compare rankings
generated by the \citeauthor{plackett1975} strengths and the points
awarded and note the degree of agreement between the two.  This
furnishes an objective assessment of the points system used.
